Learning Actionable Analytics (with applications for reducing defects and reducing runtimes)

### 2015-05-03 how.tex
Updates from ShareLaTeX

\section{Introduction}
Business  users are now demanding that data mining tools
be augmented with tools to support  business-level
interpretation of that data. For example,
at a  panel on software analytics at ICSE'12,
industrial practitioners lamented the state of the art in data mining
and software engineering~\cite{menzies12a}. Panelists commented that
``prediction is all well and good, but what about decision
making?''. That is, these panelists are more interested in the interpretations
that follow the mining, rather than just  the mining. For example:
\bi
\item Instead of just accepting  {\em predictions} on how many 
 software defects
to expect,  business users might now demand a {\em plan} to
reduce the likelihood of those defects;
\item Instead of just accepting {\em predictions} on the runtime
time of their software, business users might now demand
a {\em plan} to reduce that runtime.
\ei
In these two points, we say that a ``plan'' is a  set of proposed changes in order to better achieve some goal. 
Generating {\em plans} is a  very different task the standard
data mining tasks of {\em predicting} some discrete or numeric class.
Predictors
and classifiers only tell us what {\em is} while planners 
  tell us {\em what to change}. For the kind of business user
discussed above, planners are the preferred to mere prediction systems
since, as one gruff user once shouted at us:
\begin{quote}
{\em ``Don't tell me what \underline{is};
tell me what to \underline{change}.''}
\end{quote}
In order to satisfy the demands of this new kind of business user, we  study the generation of {\em plans} in the context
of software analytics.  
This paper presents
HOW, a novel  {\em instance-based planner} that runs in near-linear time.
In the experiments shown below, we see that HOW's plans can
significantly reduce the expected
value of:
\bi
\item The software defects found in  twelve object-oriented JAVA systems explored in 2010 by Jureczko et al.~\cite{jureczko10};
\item The runtime of the six software systems whose configurations
are explored in 2013 by Seigmund et al.~\cite{sven12};
\ei
Many papers use this above data to build {\em predictors} for defects and runtimes.
However, none of those papers then go on to use this data to build {\em planners}.

HOW's instance-based reasoning  makes its conclusions by
studying the space of examples. This is a very different approach to
{\em model-based planners} that use the model
as sub-routine within a Monte Carlo simulation~\cite{me07f} or the evolutionary 
programming methods explored by the search-based software
engineering community~\cite{krall14,harman12dec}. 
We find   HOW's instance-based approach useful since it can be applied
to domains that lack a model, but possess a historical log of observed behaviours.

Instance-based reasoning is also useful if building and certifying a model   takes a long time. Such slow-to-build models exist in software engineering. For  example, in previous work~\cite{me07f} we used models
developed by Boehm's group at the University of Southern California.
Some of those models took decades to develop and mature (from 1981~\cite{boehm81} to 2000~\cite{boehm00b}).  This is not an isolated example. One of us (Menzies) teaches a graduate class in automated model-based software engineering. 
Students in that class are asked to find (or build) a non-trivial model of
some software process (where ``non-trivial'' means something that processes
a  model at least an order of magnitude more complex than Brooks' Law). 
While the  number of such models has been growing in recent years\footnote{See the
catalogue at goo.gl/nv2AVK.}, 
the total number remains very small-- which is an observation consistent with
SE models typically require  much effort to build and certify.

@@

In summary, this paper makes five contributions:
\be
\item A general framework for augmenting a prediction system such that
it becomes a planning system;
\item A  test that recognizes  when we should {\em not}  plan: in summary, our kind of planning is
not recommended in domains that  generate poorly performing predictors;
\item A general verification rig for assessing the value of plans generated within this rig.
\item The HOW  instance-based planner.
\item A proof that HOW runs in near-linear time.
\item An experimental validation that the HOW planner is useful for reducing the
expected values of the Jureczko software defects and the Seigmund software runtimes.
\ee
The rest of this paper is structured as follows.

\section{From ``Prediction'' to ``Planning''}
A major  problem with building a planning mechanism in software engineering is ``how to verify the generated
plans?''. For example,  in the experiments shown below,  HOW comments  on how to reduce
software defects by adjusting certain static code parameters of the code\footnote{Such parameters can be mnaually adjusted by programmers or (for large scale changes) automatically adjusted using code refactoring
tools, as shown by Binkely et al. in 2006~\cite{Binkley2006} and more recently by Mkaouer et al. in 2015~\cite{Mkaouer15}.}. How are we to verify that HOW's recommendations actually work?

>>>>

\section{Introduction}
Business  users are now demanding that data mining tools
be augmented with tools to support  business-level
interpretation of that data. For example,
at a  panel on software analytics at ICSE'12,
industrial practitioners lamented the state of the art in data mining
and software engineering~\cite{menzies12a}. Panelists commented that
``prediction is all well and good, but what about decision
making?''. That is, these panelists are more interested in the interpretations
that follow the mining, rather than just  the mining. For example:
\bi
\item Instead of just accepting  {\em predictions} on how many 
 software defects
to expect,  business users might now demand a {\em plan} to
reduce the likelihood of those defects;
\item Instead of just accepting {\em predictions} on the runtime
time of their software, business users might now demand
a {\em plan} to reduce that runtime.
\ei
In these two points, we say that a ``plan'' is a  set of proposed changes in order to better achieve some goal. 
Generating {\em plans} is a  very different task the standard
data mining tasks of {\em predicting} some discrete or numeric class.
Predictors
and classifiers only tell us what {\em is} while planners 
  tell us {\em what to change}. For the kind of business user
discussed above, planners are the preferred to mere prediction systems
since, as one gruff user once shouted at us:
\begin{quote}
{\em ``Don't tell me what \underline{is};
tell me what to \underline{change}.''}
\end{quote}
In order to satisfy the demands of this new kind of business user, we  study the generation of {\em plans} in the context
of software analytics.  
This paper presents
HOW, a novel  {\em instance-based planner} that runs in near-linear time.
In the experiments shown below, we see that HOW's plans can
significantly reduce the expected
value of:
\bi
\item The software defects found in  twelve object-oriented JAVA systems explored in 2010 by Jureczko et al.~\cite{jureczko10};
\item The runtime of the six software systems whose configurations
are explored in 2013 by Seigmund et al.~\cite{sven12};
\ei
Many papers use this above data to build {\em predictors} for defects and runtimes.
However, aside from work by the authors of this paper,
none of those papers then go on to use this data to build {\em planners}.

In summary, this paper makes five contributions:
\be
\item A general framework for augmenting a prediction system such that
it becomes a planning system;
\item A  test that recognizes  when we should {\em not}  plan: in summary, our kind of planning is
not recommended in domains that  generate poorly performing predictors;
\item A general verification rig for assessing the value of plans generated within this rig.
\item The HOW  instance-based planner.
\item A proof that HOW runs in near-linear time.
\item An experimental validation that the HOW planner is useful for reducing the
expected values of the Jureczko software defects and the Seigmund software runtimes.
\ee
The rest of this paper is structured as follows.
HOW's instance-based reasoning  makes its conclusions by
studying the space of examples. As discussed in \tion{mm},
this is a very different approach to
{\em model-based planners} that use the model
as sub-routine within a Monte Carlo simulation~\cite{me07f} or the evolutionary 
programming methods explored by the search-based software
engineering community~\cite{krall14,harman12dec}.

\section{HOW: Motivation and Methods}\label{sec:mm}

 \subsectuion{Why ``HOW''?}

HOW was inspired by some recent work by Krall and Menzies and Davis on 
the GALE optimizer~\cite{krall14}. A well-known effect in machine learning (and, in computer vision
and compression~\cite{levina04}) is that the $n$ dimensions of a data set can be represented by a  much
lower-dimensional approximation. GALE finds those $m << n$ dimensions then mutates examples
towards the better end of each dimension $m_i$.

A limitation with GALE is that whenever it adjusts an instance, some model has to be called
to re-evaluate the new objective scores for the new examples. This can be a limitation
since, in some domains, no such model exists or is readily accessible.
Building and certifying a model   takes a long time. Such slow-to-build models exist in software engineering. For  example, in previous work~\cite{me07f} we used models
developed by Boehm's group at the University of Southern California.
Some of those models took decades to develop and mature (from 1981~\cite{boehm81} to 2000~\cite{boehm00b}).  This is not an isolated example. One of us (Menzies) teaches a graduate class in automated model-based software engineering. 
Students in that class are asked to find (or build) a non-trivial model of
some software process (where ``non-trivial'' means something that processes
a  model at least an order of magnitude more complex than Brooks' Law). 
While the  number of such models has been growing in recent years\footnote{See the
catalogue at goo.gl/nv2AVK.}, 
the total number remains very small-- which is an observation consistent with
SE models typically require  much effort to build and certify.

@@

Depending on the audience, we can
might call HOW a spectral learner, 
a response surface method, or  a local search operator.
{\em Spectral learners}~\cite{kamvar03} reason across $m$ dimensions synthesized
using (say) PCA~\cite{pearson1901}. For example,
algorithms like PDDP~\cite{boley98} use PCA to
recursively divide data into smaller regions where each level of recursion needs an   $O(N^2)$ analysis
to infer the principle components. HOW performs an analogous procedure using a   $O(2N)$ analysis
based on a heuristic proposed by
Faloutsos and Lin~\cite{Faloutsos1995} (see above, the ``FASTMAP'' algorithm).

{\em Response surface methods}  build a fast approximation to the function being optimized,
then uses that  approximation to take intelligent guesses about useful mutations. HOW
is a non-parametric response surface method that finds its response surface via recursive
clustering. For an example of a parametric response surface method, that uses Gaussian process
models, see Zuluaga et al.~\cite{zuluaga13}.

A {\em local search operator} that a some probability $p$, nudges  instance variables
along a dimension that improves its objective scores. HOW's recursive clustering methods
combine $n$ dimensions into $m \ll n$ so its ``nudges'' are simultaneous
adjustments along multiple dimensions. For an example of other kinds of local search algorithms, that only adjust
one variable at a time, see the literature on WALKSAT~\cite{Selman1992} and MAXWALKSAT~\cite{kautz97}.
See also the literature on local search methods in multi-objective optimization.

\be
\item Take  a population $P$ of examples
\item Separate all $p \in P$ examples into their $n$ independent variables and 
their $o$ scores;
\item Find $m \ll n$ lower dimensional approximations for the $n$ independent variables;
\item Look for the worst and best ends of each $m$  dimension (where the better ends
have better objective scores);
\item Find the $q \subseteq P$ subset of the population with ``better'' objective scores;
\item Generate the next population $P'$ by adjust the independent variables
of the $q$  examples
by a random amount (in the direction of  the  better ends of each dimension).
\item Apply some model to discover all the objective scores of $P'$
\item Set $P=P'$ and go to step 1
\ee

PCA explores the $e$
eigenvectors of the correlation matrix between $d$
variables~\cite{pearson1901}.  Usually,
only a few $e \ll d$ eigenvectors are required
since multiple inter-related variables $i,j,k \subseteq d$ can be represented
by a single  eigenvector.
Also, noisy variables from $d$ are
ignored since they  do not contribute to the signal in the data.
Further,
redundant  variables  become (approximately) parallel lines
in $e$ space. For all such redundancies \mbox{$i,j \in d$}, we
can ignore $j$
since effects that change over $j$ also
change in the same way over $i$.

Algorithms like PDDP~\cite{boley98} use PCA to
recursively divide data into smaller regions. Each
step of that process requires an $O(N^2)$ analysis
of the data to compute the components~\cite{ilin10}. Worse, it requires an evaluation of every example.
Using WHERE, competitive results can be achieved
using an $O(2N)$ analysis.
WHERE uses a heuristic proposed by
Faloutsos and Lin~\cite{Faloutsos1995}
(which
Platt has
  shown computes a Nystr\"om
  approximation to the first component of
  PCA~\cite{platt05}).
In step one, WHERE inputs
$N$
examples $N_1,N_2,..$. Secondly, WHERE
picks any
point $N_i$ at random. Thirdly, WHERE
finds the point  {\em West}~$\in N$ that is
furthest\footnote{
 For this work, we use the standard Euclidean measure recommended for
instance-based reasoning by Aha et al.~\cite{aha91};
i.e. $\sqrt{\sum_i(x_i-y_i)^2}$ where $x_i,y_i$
are values normalized 0..1 for the range min..max.}
from $N_i$.
Finally, WHERE

$m \ll n$
dimension

HOW divide training into regions where the training
data has similar properties. Each region is then examined and divided into ``west'' and ``east''
where the region's ``west'' end has examples with worst performance stores than the east
end (mnemoc: west starts with ``w'' and is short for ``worst' and east starts with ``e;;]which is shirt
for excellent).so the ``e'' in

\be
\item Clustering the training data 
\item 
(c) finding which end of the cluster has better performance scores.
(d) find

\section{Methodological Comments:\\ from ``Prediction'' to ``Planning''}
A major  problem with building a planning mechanism in software engineering is ``how to verify the generated
plans?''. For example,  in the experiments shown below,  HOW comments  on how to reduce
software defects by adjusting certain static code parameters of the code\footnote{Such parameters can be mnaually adjusted by programmers or (for large scale changes) automatically adjusted using code refactoring
tools, as shown by Binkely et al. in 2006~\cite{Binkley2006} and more recently by Mkaouer et al. in 2015~\cite{Mkaouer15}.}. How are we to verify that HOW's recommendations actually work?
